---
title: "Data Cleaning"
author: "Chinedu O"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning Steps Using R

To effectively clean data using R, use start by installing and loading the necessary packages.

***
Install useful packages using the following codes:

* install.packages("here")
* install.packages("janitor")
* install.packages("skimr")
* install.packages("dplyr")
* install.packages("tidyverse")
* install.packages("palmerpenguins")

Once the installation is completed, load the library in R studio using the following codes:


```{r message=FALSE}
library(here)
library(janitor)
library(skimr)
library(dplyr)
library(tidyverse)
```

For this cleaning, we used the Palmerpenguins dataset pre-installed

```{r message=FALSE}
library(palmerpenguins)
```

## Get an overview of your data set
### skim_without_charts Function
To get a summarized overview of your data set, you can use different codes that provide different perspectives on our data.  First, we will use the skim_without_charts function to get a summary as shown below.
*NB: This function is handy because it gives us an overview of the missing values and helps us to determine how we want to handle them. Later we will see a code for dropping missing values [drop_na()]

```{r message=FALSE}
skim_without_charts(penguins)
```

### glimpse Function
We can also use the function glimpse to obtain a similar result. However, this is shorter, more concise, and does not provide the statistics of the missing value. Although it gives us a quick peep of the variables in our data set and their types as seen in the result below.

```{r message=FALSE}
glimpse(penguins)
```
### head Function
The head function is used to get a quick overview of our data frame as it is arranged. It shows the first six rows of the data.

```{r message=FALSE}
head(penguins)
```

## Data manipulation
#### Filtering with R
To filter our data, we can use select to choose a specific column(s).   For example, let's say we wanted to see the total number of different penguin species, we can achieve these by:
NB: In the code below, %>% is a pipe, it allows us to execute different operations with the variable as if it is flowing in a pipe.

```{r message=FALSE}
population_by_island <- penguins %>% 
  select(species) %>% 
  count(species)
```

In the above code chunk, we stored the filtered data in a new data frame population_by_island. We can view it by using the View function or by calling the variable.

```{r message=FALSE}
population_by_island 
```



#### Changing a column name
First, we can view all the column names using the colnames function and change the name using the rename function. The syntax is written below

```{r message=FALSE}
colnames(population_by_island)
```
```{r}
population_by_island <- population_by_island %>% 
  rename(total_population=n)
population_by_island
```

We can also achieve standardization of code by making the column names uniformly upper case or lower case to achieve that we use the following syntax

```{r message=FALSE}
rename_with(population_by_island, toupper)
```
 Then, back to lowercase
 
```{r message=FALSE}
rename_with(population_by_island, tolower)
```

We can additionally use the clean_names function to ensure the column name contains only strings, numbers, or underscore and without prohibited characters

```{r message=FALSE}
clean_names(population_by_island)
```



#### Sorting, filtering and Arranging 
We can sort and arrange data using arrange, filter, or sort functions. For example. we can sort the population_by_island table in ascending order of population using this chunk of code:

```{r message=FALSE}
population_by_island %>% 
  arrange(total_population)
```

Or in alphabetical order by species 
```{r message=FALSE}
population_by_island %>% 
  arrange(species)
```
To sort in descending order we introduce a minus sign "-" before the column name or use the desc function. Look at the example below


```{r message=FALSE}
population_by_island %>% 
  arrange(desc(species))
# The code chunk below will achieve the same result.
# population_by_island %>% 
  #arrange(-species)
```


#### Groupping
To group data, we use the group_by function. We can group data and summarize by group. See sample code below.
  *NB: The drop_na is used to drop columns with na.*

```{r message=FALSE}
penguins %>% 
  group_by(island) %>% 
  drop_na() %>% 
  summarise(mean_bill_length_mm = mean(bill_length_mm), 
            max_bill_length_mm = max(bill_length_mm),
            min_bill_length_mm = min(bill_length_mm))
```


### Other Functions for Data Cleaning
We have the unite function and separate function which as their names imply helps us to join two columns or split one column into two. Let's see an example.

#### Creating a data frame
First, we create a data frame containing surname, first name, and department and we will join the surname and first name to form full name. 

```{r message=FALSE}
surname <- c("Asha", "Water", "Levi", "Agu", "Isike")
firstname <- c("Stanley", "Linus", "Pulna", "Victor", "Anna")
department <- c("HR", "Tech", "IT", "CS", "MG")

office_df <- data.frame(surname, firstname, department)
office_df
```
#### Merging Columns
Next, we will use the unite function to create a new column called full_name which concatenates the surname and first name separated by a space.

```{r message=FALSE}
office_v2 = unite(office_df, "full_name", surname, firstname,sep = " ")
office_v2
```

#### Splitting Columns
From the above code, we saved the resulting table into office_v2. Next, we use the separate function to split the full name back into surname and first name.

```{r message=FALSE}
separate(office_v2, full_name, c("surname", "firstname"), sep=" ")
```


#### Add a new column to data
Let's assume we have a vector salary that contains the salaries of the employees and we want to add it to the table, we can do so using the mutate function as shown below.

```{r message=FALSE}
salaries <- c(6500, 7000, 7300, 7500, 9000)
mutate(office_df, salary = salaries)
```
#### Data manipulation in new columns
We can also perform a little arithmetic calculation by putting the tax net salaries into the mutate function as follows
First, let's assume they pay 23% tax and their net is 77% of what they earn. we can express that in the mutate function as follows:


```{r message=FALSE}
salaries <- c(6500, 7000, 7300, 7500, 9000)
mutate(office_df, salary = salaries, tax = salaries*.23, net = salaries*.77)
```

With this, we have come to the end of data cleaning and manipulation. 

Checkout my next work on data visualization to learn how to obtain a plot like this:

```{r pressure, echo=FALSE}
plot(pressure)
```



>**Stay motivated!!** 
